---
title: "TC: Amplicon analysis"
author: "Maria Rebolleda Gómez"
date: "01/10/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---


Reads from Argonne came in a single file. To demultiplex I used [idemp](https://github.com/yhwu/idemp) and then change their names for simpler calling. 
```{bash}
~/Software/idemp/idemp -b barcodes_good2.txt -I1 Run1_I1_001.fastq -R1 Run1_R1_001.fastq -R2 Run1_R2_001.fastq -o demultiplexed2
~/Software/idemp/idemp -b barcodes_good.txt -I1 Run2_I1_001.fastq -R1 Run2_R1_001.fastq -R2 Run2_R2_001.fastq -o demultiplexed2
mkdir unsigned
mv *_unsigned* unsigned

```

Filenames start with `Run$N_R$1_001.fastq` and then after `_` their label id. Change the name to something simpler
```{bash}
for file in *"_001.fastq_"*; do mv -- "$file" "${file//_001.fastq/}"; done

```
Demultiplexed samples are available in NCBI SRA (Accession: PRJNA992630).

```{bash}
#Check quality of data with fastq
mkdir 02_fastQC
fastqc *.fastq.gz --outdir="/media/mariargz/Extra Drive 1/TC_AmpliconSequencing/demultiplexed2/02_fastQC"
```


# DADA2 Pipeline
Using demultiplexed files. 

## Load Packages and Prep Data
```{r}

# Install and load in dada2 package
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("dada2", version = "3.17")

library(dada2)
library(data.table)
library(tidyverse)


# Assign path to data (change to the right path to directory)
#parent = "Path to directory /CommmunityAssembly_16s"

# Assign directory to store ouput from analyses
path.out <- paste0(parent, "/Output/")
# Assign directory to store ESV data 
path.data<- paste0(parent,"/Processed_Data/")
# Assign directory to sequences (NCBI SRA; Accession: PRJNA992630)
# path.in<- "Directory to sequences"

# Set working directory
setwd(path.out)


# Create address to access fastq files
a <- list.files(path.in)

# Create list of file addresses for each fastq file
fnFs <- sort(list.files(path.in, pattern="R1.TC", full.names = TRUE))
fnRs <- sort(list.files(path.in, pattern="R2.TC", full.names = TRUE))

# Obtain information of the sample (between periods). 
sample.names <- basename(fnFs) %>% strsplit("\\.") %>%
  sapply(`[`,2)


```

## Read Quality Profiles
```{r}
#Select a random set of samples to look at
rdm <- sample(1:length(fnFs), 25)
# Forward
plotQualityProfile(fnFs[rdm])
# Reverse
plotQualityProfile(fnRs[rdm])

```

Quality drops ~200bp in both forward and reverse. Some reads are low quality (minority in each sample)

## Filter and Trim
```{r}
# Place filtered files in filtered/subdirectory
filtFs <- file.path(path.in, "03_filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path.in, "03_filtered", paste0(sample.names, "_R_filt.fastq.gz"))

# Filtering: Using parameters for our data

out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(230,230),
                     maxN=0, maxEE=c(1,2), minQ=11, rm.phix=TRUE,
                     compress=TRUE, multithread=FALSE) 

names(filtFs) <- sample.names
names(filtRs) <- sample.names

# This step is very slow so it makes sense to save this intermediate step
#saveRDS(out, "01-out.rds")
```


## Learn Error Rates
```{r}
set.seed(20)
#learn the error rates for forward reads
errF <- learnErrors(filtFs, nbases=1e8, multithread=TRUE)
#learn the error rates for reverse reads
errR <- learnErrors(filtRs, nbases=1e8, multithread=TRUE)
# plot error rates
plotErrors(errR, nominalQ=TRUE)
plotErrors(errF, nominalQ=TRUE)
```

## Dereplicate, and merge
```{r}
#Creat empty list to save the merged reads. 
mergers <- vector("list", length(sample.names))
names(mergers) <- sample.names

for(sam in sample.names) {
  derepFs <- derepFastq(filtFs[[sam]])
  dadaFs <- dada(derepFs, err=errF, multithread=TRUE)
  derepRs <- derepFastq(filtRs[[sam]])
  dadaRs <- dada(derepRs, err=errR, multithread=TRUE)
  merger <- mergePairs(dadaFs, derepFs, dadaRs, derepRs)
  mergers[[sam]] <- merger
}

rm(derepFs); rm(derepRs)

#Make and save sequence table
seqtab <- makeSequenceTable(mergers)
#saveRDS(seqtab, "02-seqtab.rds") 

dim(seqtab)

#Check sequence lengths (must be around 250bp)
hist(nchar(getSequences(seqtab)))

#Percentage of sequences longer than 256 --> 0.17%
sum(nchar(getSequences(seqtab))>256)/dim(seqtab)[2]*100
##Percentage of sequences shorter than 248 --> 0
sum(nchar(getSequences(seqtab))<248)/dim(seqtab)[2]*100

# Note: More stringent filtering reduced the amount of weird size fragments

#Remove sequences much shorter or longer than expected
seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% seq(248,256)]

```

## Merge paired reads, sequence table, remove chimeras
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab2, method="consensus", multithread=TRUE, verbose=TRUE)
#saveRDS(seqtab.nochim, "seqtab.nochim.rds")
dim(seqtab.nochim)

#Save seqtab
write.csv(seqtab.nochim, paste0(path.data,"ESVtable.csv"))
```

### Track Reads Through Pipeline
```{r}
### Table
getN <- function(x) sum(getUniques(x))
track <- cbind(out,  sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "merged", "nonchim")
rownames(track) <- sample.names
track
track.df <- as.data.frame(track)


#### Plot ####

# Plot the number of reads through the steps in the dada2 pipeline
library(data.table)
setDT(track.df, keep.rownames = TRUE)[]
colnames(track.df)[1] <- "Sample.ID"
track.df <- as.data.frame(track.df)
track.names <- names(track.df)

summary(track.df)
reads_tot <- apply(track.df[2:5],2,sum)
percents <- reads_tot[2:4]/reads_tot[1]

pdf("Plots/reads_tracking.pdf")
matplot(t(track.df[, -1]), type="l", ylab = "Number of Reads", xlab = "Step", xaxt = 'n')
axis(1, at=1:(length(track.names)-1), labels=track.names[2:length(track.names)])
dev.off()

water <- grep("H2O",track.df$Sample.ID)
track.df[water,]

hist(track.df$nonchim, main = "Number of reads per sample", xlab = "Number of reads")
min_reads <- min(track.df$nonchim) #8064!! - This is a great coverage, so I can rarefy at this level. 
```


## Assign Taxonomy
```{r}
#### Assign Taxonomy using SILVA database ####
taxa <- assignTaxonomy(seqtab.nochim, "/media/mariargz/Extra Drive 1/MimulusMicrobiome/raw/silva_nr_v132_train_set.fa.gz", multithread=TRUE)
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)

write.table(taxa, file= paste0(path.data,"taxa_silva_v132.txt"))

```

# Process OTU and evaluate relative abundances
## Load new packages
```{r}
#BiocManager::install("phyloseq")
library(vegan)
```

## Rarefaction and filtering of ESV table
```{r}
# Rarefaction to the minimum number of reads in a sample
rarefied_data <- rrarefy(seqtab.nochim, min_reads) 

#Check dimensions of data 
dim(rarefied_data)
#[1] 480 901 (samples ESVs)

# Remove ESVs that are empty
colsums <- colSums(rarefied_data)>0
rarefied_noempty <- rarefied_data[,colsums]

#Check dimensions of data again to make sure empty ESVs were removed
dim(rarefied_noempty)
#[1] 480 593 

write.csv(rarefied_noempty, paste0(path.data, "rarefied_data.csv"))
```

